---
title: "Relative_Transitions"
author: "Frieda"
format: html
editor: visual
---

# 0. Required Libraries

```{r}
#| label: libraries
# Clean environment:
rm(list=ls())

# Libraries:

library(evobiR) # reorder data by tree

library(phytools) # v. ‘1.9.16’ for phylo stuff
library(dplyr) # for data handling
library(evobiR) # v. ‘1.1’ To reorder data based on tree
library(geiger) # v. ‘2.0.11’ for Mk Models
library(tictoc) # for time measurements of computations
library(tidyr) # data handling
library(ggplot2) 
library(prettyGraphs) # for transparancy settings while plotting

```

##### 1. & 2. in Script: 1_CumulativeRates.qmd

# 

## 3. Relative Transition rates

# A) Legumes

this part runs a while.. the loop takes about 22 minutes.

## Adaptation of ctt function (in loop, save to list)

Reference:
[ctt.R](https://github.com/liamrevell/phytools/blob/master/R/ctt.R)
(Phytools package)

```         
> packageVersion("phytools") [1] ‘1.9.16’
```

```{r}
#| label: loop
#| eval: false

rm(list=ls())
mtrees <- readRDS("../output/RDS/legumes_mtrees_multiSimmap.rds")
tree <- readRDS("../output/RDS/Legumes_MCC.rds")
### Required function ========================================================================== #
getChanges<-function(tree){
  states<-sort(unique(getStates(tree)))
  nc<-sapply(tree$maps,length)-1
  b<-which(nc>0)
  nc<-nc[b]
  xx<-vector()
  H<-nodeHeights(tree)
  for(i in 1:length(b)){
    for(j in 1:nc[i]){
      ss<-names(tree$maps[[b[i]]])[j+1]
      x<-rep(H[b[i],1]+cumsum(tree$maps[[b[i]]])[j],2)
      xx<-c(xx,setNames(x[1],
                        paste(names(tree$maps[[b[i]]])[j:(j+1)],
                              collapse="->")))
    }
  }
  xx
}



# ============================================================================================= #
# Create a function to find the segment ID for a given time value:
# ============================================================================================= #

find_segment_id <- function(time) {
  segment_indices <- cut(time, breaks = segs_df[, 2], labels = FALSE)
  return(segs_df$segsID[segment_indices])
}

## ============================================================================================= #

```

```{r, include=FALSE, eval = F}
#### Subset to test script::
set.seed(123)
species_samp <- sample(tree$tip.label, 30)

tree <- keep.tip(tree, species_samp)
mtrees <-keep.tip(mtrees, species_samp)

######
```

```{r}
tic( "loop")

# ================================================================= #
# Working the multiSiammp
# ================================================================= #

# Get all changes from all simulations so that we can loop through them:
changes <- sapply(mtrees, getChanges) 
  
  
# Transform the original changes list of matrices from sapply() for all simulations to dataframe 
  # add trans_no to each transition based on grouping by transition 
  # (i.e, there are two trans_no = 1, one for each transition type)
  
  
i_changes_list <- list() 
for (i in 1:length(changes)) {
  i_changes <- changes[[i]] # for each simulation from multiSimmap
  
  # extract transition types and times and assign ID column for simmap simulation
  i_changes_df <- data.frame(transition = names(i_changes), time = i_changes, N_sim = i)
  i_changes_df$transition <- as.factor(i_changes_df$transition)
    
  # Add trans_no to each transition from a group (transition type)
  i_changes_df <- i_changes_df %>%  arrange(time) %>%
    group_by(transition) %>%
    mutate(trans_no = row_number())
    
  i_changes_list[[i]] <- i_changes_df
}
  
# Final Table without edge.lengths:
changes_table <- plyr::rbind.fill(i_changes_list, fill=T)
changes_table_wo_edge <- changes_table




# ====================================================================================================== #
# ~ Step 1: Loop over each simulation and extract 
# - breaks for segments
# - edge.lengths
# - rates for each character state transition
# ====================================================================================================== #

object_list <- list()
rates_list <- list()

# First Level of the loop  
for (Nsimulation in 1:100){ 
  
  # ====================================================================================================== #
  # Setting global variables for the Analysis
  # ====================================================================================================== #
  
  # set the number of time-segments for which everything will be calculated:
  b <- 20    
  
  # reduce to one tree to extract parameters from each simulation separately:
  tree <- as.phylo(mtrees[[Nsimulation]])
  
  # extract time since root
  h <- max(nodeHeights(tree))
  
  # calculate the time segments based on time since the origin and the time segments that were set by the user
  segs <- cbind(seq(0,h-h/b,h/b),
                seq(1/b*h,h,h/b))

  
  # ====================================================================================================== #
  # Transitions & Timing  (getChanges)
  # ====================================================================================================== #
  
  
  # ================================================================= #
  # Working changes from each simmap simulation separately:
  # ----------------------------------------------------------------- #
  # Changes per Segments:
  # ================================================================= #
  


  # Loop through all changes & Calculate average number of changes per time interval (segs; set with "b")
  
  nchanges <- rep(0,b)
  for(i in 1:length(changes)) { # loop through all simulations from simmap  

    # loop through all individual changes from one simmap simulation and calculate relative changes per time segment
    # nchanges = collects the relative number of changes (of all changes) per segment
    
    for(j in 1:length(changes[[i]])) { 
      ind<-which((changes[[i]][j]>segs[,1])+ 
                 (changes[[i]][j]<=segs[,2])==2)
      nchanges[ind]<-nchanges[ind]+1/length(changes)
      
      } # closing of 2nd level
    } # closing of 1st level

  rm(i,j)  
  
  # ================================================================= #
  # Working changes from each simmap simulation separately:
  # ----------------------------------------------------------------- #
  # Edge Lengths per Segments:
  # ================================================================= #

  # Edge.Lengths = Number of species in the phylogeny per time slot (increases naturally with time)
  # ! Note: has to be accounted for when showing transition rates
  # Here we calculate edge lengths for each segement separatly:
  
  # -------------------------------------------------------------------------------------------------------- #
  

  
  # Compute Lineages-Through-Time with ltt() function from the phytools package to extract edge.lengths from
  LTT <- ltt(tree,plot=FALSE, log.lineages = F)
  cols<-setNames(palette()[c(2,4,1)],c(unique(names(changes[[1]])),
    "total"))

  plot(LTT,col=cols,bty="n",las=1,lwd=4,show.tree=FALSE,ylim=c(1,Ntip(tree)),
    xlab="time (above the root)")
  
  
	# transform LTT data so that we can use it:
  LTT <- cbind(LTT$ltt[2:(length(LTT$ltt)-1)],            # Number of species at each node minus the root
               LTT$times[2:(length(LTT$ltt)-1)],          # "start" time for each node minus the root
               LTT$times[3:length(LTT$ltt)])              # "end" time for each node minus the root
  
  # -------------------------------------------------------------------------------------------------#
  
  ## Now we loop through all segments to extract matching edge.lengths =================================== #
  
  # set count variable:
	ii<-1

	# Create an empty vector to store edge lengths
	edge.length <- rep(0, nrow(segs))

	
	# 2nd level loop: 
	# (for one tree because LTT is calculated for one tree)
   for (i_segs in 1:nrow(segs)) {
     

     # Find the indices of LTT segments that overlap with the current segs segment
     overlap_indices <- which(LTT[, 2] <= segs[i_segs, 2] & LTT[, 3] >= segs[i_segs, 1])
  

     # Calculate edge length for the current segs segment using vectorized operations
     # 3rd level loop: 
     for (ii in overlap_indices) {
     
       # edge.length should contain the edge lengths for all segments:  
       edge.length[i_segs] <- edge.length[i_segs] +        
         LTT[ii, 1] * # counts number of lineages between time intervals
         (min(segs[i_segs, 2], LTT[ii, 3]) - # end time interval
            max(segs[i_segs, 1], LTT[ii, 2]))  # start time interval

     
       } # 3rd level closing
     
   } # 2nd level closing
	
	
	edge.length_backup <- edge.length
	edge.length <-c(1, edge.length)
  # edge.length should contain the edge lengths for all segments in a given simulation
 	
	# adding segsID coulmn to edge lengths:
	edge.length_df <- data.frame(edge.length = edge.length, segsID = seq(1,b+1))
	
	

	# ================================================================= #
  # Working the multiSimmap:
  # ----------------------------------------------------------------- #
  # Edge Lengths at each transition event
  # ================================================================= #
	## Adapted code from the segments (above) by Frieda
	
	# we need the table for all simmaps: changes_table so that we can loop through it
	
	# set count variable:
	ii<-1

	# Create an empty vector to store edge lengths
	edge.length_transitions <- rep(0, nrow(changes_table))
  edge.lengths_all <- list()
	changes_times <- changes_table$time
	
	
	# 2nd level loop: 
	# (for one tree because LTT is calculated for one tree)
   for (i_changes in 1:nrow(changes_table)) {

     # Find the indices of LTT segments that overlap with the current segs segment
     
     # "which start times of the Lineages are smaller or equal to the time the change happened and which end times of the lineages are larger or equal to the time the change happend."
     # --> this gives the node numbers that were there when the transition happend
     
     overlap_indices <- which(LTT[, 2] <= changes_times[i_changes] & LTT[, 3] >= changes_times[i_changes])
  

     # Calculate edge length for the current segs segment using vectorized operations
     # 3rd level loop: 
     
     for (ii in overlap_indices) { # ii = number of lineages when a transition occurred
     
       # edge.length should contain the edge lengths for all segments:  
       edge.length_transitions[i_changes] <- edge.length_transitions[i_changes] +        
         LTT[ii, 1] 
       
       edge.length_transitions2 <- data.frame(time = changes_times[i_changes], edge.length = edge.length_transitions[i_changes])

     
       edge.lengths_all[[i_changes]] <- edge.length_transitions2
       
       } # 3rd level closing
     
   } # 2nd level closing
	
  
  edge.lengths_all_df <- plyr::rbind.fill(edge.lengths_all)
  changes_table_w_edge <- merge(changes_table_wo_edge, edge.lengths_all_df, by = "time")
  # Write to file (once is enough)
  #write.csv(changes_table_w_edge, "../output/CSV/Legumes_changes_table_w_edgelengths.csv")
	

# ============================================================================================= #
# Create a dataframe with transitions, times, edge.lengths, segment IDs.
# ============================================================================================= #
	  
  ## We are here = first level loop: along simulations
	  
	 	# Again extract changes for each simulation separately:
	  current_simulation <- as.data.frame(changes[[Nsimulation]], 
	                                      row.names = names(changes[[Nsimulation]]))
	  names(current_simulation) <- c("time")
	  current_simulation$transition <- rownames(current_simulation)
	 	current_simulation$N_sim <- Nsimulation
	 	current_simulation <- current_simulation %>% arrange(time)
	 	current_simulation$IDtransition <- seq(1:nrow(current_simulation))

	 	
	 	# ============================================================================================= #
	 	# Working the segments: 
	 	# ============================================================================================= #

	 	
	 	# Create a vector of segment IDs
	 	segs_df <- as.data.frame(segs)
	 	new_row <- data.frame(V1 = 0, V2 = 0) # add a row that starts from 0 and ends with 0.. we need this later on for plotting
	 	segs_df <- rbind(new_row, segs_df)
	 	segs_df$segsID <- seq(1,b+1) # now we can calculate the breaks from it

	 	# Use the apply function to find the segment ID for each row in current_simulation
	 	current_simulation$segsID <- apply(current_simulation, 1, function(row) {
  
	 	  time <- as.numeric(row["time"])
	 	  return(find_segment_id(time))

	 	  }) # closing loop from within the apply-function (2nd level loop closing)

	
	 	# assign edge.length based on segment ID
	 	current_simulation <- unique(merge(current_simulation, edge.length_df, by="segsID", all = T))
	 	names(segs_df) <- c("segs_start", "segs_end", "segsID")
	  current_simulation <-	unique(merge(current_simulation, segs_df, by = "segsID", all = T))
	 	
	  # Calculate rate for each simulation:
	 	current_simulation_rates <- current_simulation %>% 
	 	  group_by(segsID, N_sim, transition) %>% 
	 	  mutate(Nchanges = length(!is.na(IDtransition))) %>% 
	 	  mutate(rate = Nchanges / edge.length)

	 rates_list[[Nsimulation]] <- current_simulation_rates # we need the rates_list object for plotting (!)
	  
	 object <-list(segments=segs, 
	               nchanges=nchanges, 
	               edge.length=edge.length[-1], 
	               tree=tree)
	 
	class(object)<-"ctt"
	object_list[[Nsimulation]] <- object 
	  
} # 1st level closing

toc()  # 1218.6 sec elapsed = 20.3 min

class(object_list) <-"multiCtt"

# -------------------------------------------------------------------------------------- #

# Make dataframe with rates from the rates_list object for all simulations from the simmap:
rates_df <- plyr::rbind.fill(rates_list)


# Now set columns: rates, Nchanges to 0 if there was no transition
# Calculate the average transition time based on numbered transitions from 100 simmap simulations
# set average timing of 0-transition to mean-segment time 
# then order the dataframe by segment ID:
rates_df <- rates_df %>%
  group_by(N_sim, segsID, transition) %>%
  
  mutate(rate = ifelse(is.na(transition), 0, rate), # set transition rates 0 if there was no transition
         
         Nchanges = ifelse(is.na(transition), 0, Nchanges), # set Nchanges 0 if there was no transition
         
         segs_mean = (segs_end+ segs_start)/2) %>% # Adjust the threshold (1e-6) as needed
  
  mutate(avg_time = ifelse(Nchanges == 1, time, mean(time))) %>%  # set average time = time if there was only 1 change
  
  mutate(avg_time = ifelse(is.na(avg_time), segs_mean, avg_time)) %>% # set time to mean time bin if there was no transition
  
  arrange(segsID) %>% unique() 


#save.image("~/GitHub/Evo-Legumes-Herbivory/LegumesArmature/output/RData/Legumes_rates_ws.RData")
```

This now gives us the object "object_list", which can be used inside the
phytools function plot.multiCtt() to get the overall increases and
decreases for the rates (not separated by each rate). We can also plot
the rates of each simulation separately using plot.ctt() from phytools
Also, we have just created the rates_df for plotting

```{r}
load("../output/RData/Legumes_rates_ws.RData")
```

## plot.multiCtt()

```{r}
#| label: plot.multiCtt() function Legumes

plot.multiCtt<-function(x,...){
  if(hasArg(alpha)) alpha<-list(...)$alpha
  else alpha<-0.05
  segments<-x[[1]]$segments
  nchanges<-sapply(x,function(x) x$nchanges)
  if(hasArg(type)) type<-list(...)$type
  else type<-"rate"
  edge.length<-sapply(x,function(x) x$edge.length)
  obj<-list(segments=segments,nchanges=rowMeans(nchanges),
            edge.length=rowMeans(edge.length),tree=x[[1]]$tree)
  class(obj)<-"ctt"
  lower<-max(floor(alpha/2*length(x)),1)
  upper<-min(ceiling((1-alpha/2)*length(x)),ncol(nchanges))
  xx<-max(nodeHeights(x[[1]]$tree))-as.vector(t(segments))
  xx<-c(xx,xx[length(xx):1])
  y.lower<-if(type=="number") apply(nchanges,1,sort)[lower,] else
    if(type=="rate") apply(nchanges/edge.length,1,sort)[lower,]
  y.upper<-if(type=="number") apply(nchanges,1,sort)[upper,] else
    if(type=="rate") apply(nchanges/edge.length,1,sort)[upper,]
  y.lower<-as.vector(rbind(y.lower,y.lower))
  y.upper<-as.vector(rbind(y.upper,y.upper))
  yy<-c(y.lower,y.upper[length(y.upper):1])
  args<-list(...)
  if(!is.null(args$alpha)) args$alpha<-NULL
  if(is.null(args$col)) args$col<-"blue"
  if(is.null(args$ylim)) args$ylim<-range(yy)
  args$x<-obj
  do.call(plot,args)
  polygon(xx,yy,col=make.transparent("grey",0.4),border=0)
}

plot.multiCtt(object_list, type="rate")
plot.multiCtt(object_list, type="number")
```

We can also plot the rates of each simulation separately using
plot.ctt() from phytools:

```{r}
#|label: plot.ctt() function

plot.ctt<-function(x,...){
  h<-max(nodeHeights(x$tree))
  args<-list(...)
  if(!is.null(args$type)){ 
    type<-args$type
    args$type<-NULL
  } else type<-"rate"
  if(!is.null(args$show.tree)){
    show.tree<-args$show.tree
    args$show.tree<-NULL
  } else show.tree<-FALSE
  if(!is.null(args$add)){ 
    add<-args$add
    args$add<-NULL
  } else add<-FALSE
  if(is.null(args$ylim)) 
    args$ylim<-if(type=="number")c(0,max(x$nchanges)) else 
      c(0,max(x$nchanges/x$edge.length))
  if(is.null(args$xlim))
    args$xlim<-c(max(x$segments),min(x$segments))
  if(is.null(args$lwd)) args$lwd<-2
  if(is.null(args$xlab)) args$xlab<-"time since the present"
  if(is.null(args$ylab)) 
    args$ylab<-if(type=="number") "mean number of changes" else
      "mean number of changes / total edge length"
  args$type<-"l"
  args$x<-h-as.vector(t(x$segments))
  args$y<-if(type=="number") rbind(x$nchanges,x$nchanges) else 
    rbind(x$nchanges/x$edge.length,x$nchanges/x$edge.length)
  if(!add) do.call(plot,args)
  else do.call(lines,args)
  if(show.tree) plotTree(x$tree,add=TRUE,ftype="off",lwd=1,
                         color=make.transparent("blue",0.1),mar=par()$mar,
                         direction="leftwards",xlim=args$xlim)
}


plot.ctt(object_list[[1]], type="rate")
plot.ctt(object_list[[1]], type="number")
```

# 2.1. Post-processing of data

***To do:***

Currently, if there was no transition of a specific type at a specific
time, there is no indication of a rate. However, the rate should be zero
if there was no transition. In the following code I try to replicate
each time segment twice and add transitions for both types with rates ==
0.\
I'm not done with this yet.

```{r}

# Replace NAs in Transition types to set the rates to 0 where no transition occurred:
rates_df_copy <- rates_df

# Replace "NA" with the first transition type
rates_df_copy$transition[is.na(rates_df_copy$transition)] <- unique(rates_df_copy$transition)[[2]]

# Create a second copy of the data frame with "NA" transitions
rates_df_copy2 <- rates_df

# Replace "NA" with the second transition type
rates_df_copy2$transition[is.na(rates_df_copy2$transition)] <- unique(rates_df_copy2$transition)[[3]]

# Combine the two copies to get the final data frame with replicates
final_df <- bind_rows(rates_df_copy, rates_df_copy2)


########## ====================================================================================== ####
# Specify the transition type for which you want to ensure rates
specific_transition_type <- unique(rates_df_copy$transition)[[2]]

# Generate a data frame with all possible combinations of segsID and the specific transition type
all_combinations <- expand_grid(
  segsID = unique(final_df$segsID),
  transition = specific_transition_type
)

# Left join the generated data frame with the original data frame
final_df1 <- all_combinations %>%
  left_join(final_df, by = c("segsID", "transition")) %>%
  mutate(rate = ifelse(is.na(rate), 0, rate))# Generate a data frame with all possible combinations of segsID and the specific transition type
all_combinations <- expand_grid(
  segsID = unique(final_df1$segsID),
  transition = specific_transition_type
)

# Left join the generated data frame with the original data frame
final_df1 <- all_combinations %>%
  left_join(final_df1, by = c("segsID", "transition")) %>%
  mutate(rate = ifelse(is.na(rate), 0, rate))
########## ====================================================================================== ####
# Specify the transition type for which you want to ensure rates
specific_transition_type <- unique(rates_df_copy$transition)[[1]]

# Generate a data frame with all possible combinations of segsID and the specific transition type
all_combinations2 <- expand_grid(
  segsID = unique(final_df$segsID),
  transition = specific_transition_type
)

# Left join the generated data frame with the original data frame
final_df2 <- all_combinations2 %>%
  left_join(final_df, by = c("segsID", "transition")) %>%
  mutate(rate = ifelse(is.na(rate), 0, rate))# Generate a data frame with all possible combinations of segsID and the specific transition type

all_combinations2 <- expand_grid(
  segsID = unique(final_df2$segsID),
  transition = specific_transition_type
)

# Left join the generated data frame with the original data frame
final_df2 <- all_combinations2 %>%
  left_join(final_df2, by = c("segsID", "transition")) %>%
  mutate(rate = ifelse(is.na(rate), 0, rate))
########## ====================================================================================== ####



# Sort the data frame if needed
final_df <- final_df %>% arrange(segsID)  # You can specify the sorting order

# Now final_df contains rows with "NA" in the "transition" column replaced by two different transition types.










ggplot(final_df, aes(y = rate, x = segs_end, fill = factor(transition))) +
  geom_col(position = "identity")+
  ylim(0, 0.015)+
  theme_light()

```

### Plotting the rates

#### My adaptation of pollination study using

```{r}
## below adapted from other script:  https://github.com/rubysaltbush/pollination-macroevolution/blob/main/scripts/analysis/simmap.R


armature_transitions <- rates_df
armature_transitions$transition <- as.factor(armature_transitions$transition)


# build new data frame with cumulative number of transitions
armature_trans_cumul <- data.frame()
for(n in 1:100){
  trans <- armature_transitions %>% ungroup() %>% 
    dplyr::filter(N_sim == n) %>%
    dplyr::group_by(transition) %>%arrange(time) %>%
    dplyr::mutate(trans_no = row_number(N_sim))
  
  armature_trans_cumul <- rbind(armature_trans_cumul, trans)
}
rm(n, trans)




# =========================================================================================================== #
#  Now we are summarizing over the 100 simulations:
# =========================================================================================================== #
##### copied from somewhere else in the code from the same script....:


# first need to know average number of transitions and across simulations, rounded. 
trans_avg_length <- armature_trans_cumul %>%
  dplyr::group_by(transition, N_sim) %>%
  dplyr::mutate(no_trans = max(trans_no)) %>%
  dplyr::ungroup() %>%
  dplyr::filter(trans_no == no_trans) %>%
  dplyr::group_by(transition) %>%
  dplyr::mutate(avg_length = round(mean(no_trans), digits = 0)) %>%
  dplyr::select(transition, avg_length) %>%
  dplyr::ungroup() %>%
  dplyr::distinct()

# now knowing this, can rearrange data and average times and edge.lengths across rows
avg_trans_times <- armature_trans_cumul %>%
  dplyr::group_by(transition, trans_no) %>%
  dplyr::mutate(avg_time = mean(time)) %>%
  dplyr::mutate(SE_time = sqrt(var(time) / length(time)),
                avg_edge.length = mean(edge.length)) %>% ungroup() %>%
  group_by(N_sim, segsID, transition) %>%
  mutate(avg_Nchanges = mean(Nchanges)) %>%
  dplyr::select(transition, trans_no, avg_time, SE_time, avg_edge.length, avg_Nchanges) %>%
  dplyr::distinct()

# reduce avg_trans_times to trans_avg_length
avg_trans_times_a2na <- avg_trans_times %>%
  dplyr::filter(transition == "armature->no_armature") %>%
  dplyr::filter(trans_no <= trans_avg_length[1,2])
avg_trans_times_na2a <- avg_trans_times %>%
  dplyr::filter(transition == "no_armature->armature") %>%
  dplyr::filter(trans_no <= trans_avg_length[2,2])
avg_trans_times <- rbind(avg_trans_times_a2na, avg_trans_times_na2a)
rm(avg_trans_times_a2na, avg_trans_times_na2a)




#### Plotting ======================================

# install.packages("prettyGraphs") #to set transparancy of colors with "alpha"
library(prettyGraphs)

myColours = c("lightgrey", "#FFBB00")
myColoursAlpha <- add.alpha(myColours, alpha=0.4)
my_cols <- setNames(myColoursAlpha, c("armature->no_armature", "no_armature->armature"))


```

#### same thing in ggplot2 (rates plot)

```{r}
library(ggplot2)


plot.ctt(object, type="rate")

no_arm_to_arm <- armature_transitions %>%
      filter(transition == "no_armature->armature")
    
arm_to_no_arm <- armature_transitions %>%
      filter(transition == "armature->no_armature")


ggplot() +  
  # Armature -> No Armature:
      geom_point(data = arm_to_no_arm, aes(x = time, y = rate, colour = "Armature -> no Armature"), pch = 15, alpha = 0.3) +
   # No Armature -> Armature:
      geom_point(data = no_arm_to_arm, aes(x = time, y = rate, colour = "No Armature -> Armature"), pch = 17, alpha = 0.3) +
      geom_line(data = arm_to_no_arm, aes(x = time, y = rate, colour = "Armature -> no Armature", group = N_sim)) +
      geom_line(data = no_arm_to_arm, aes(x = time, y = rate, colour = "No Armature -> Armature",group = N_sim)) +
    # Set plot aesthetics
    theme_light() + 
    xlim(45, 0) +
    ylim(0, 0.23) +
  scale_color_manual( values=c("lightgrey", "#FFBB00"))+
    labs(
      x = "Time since the origin of clade (mya)",
      y = "Relative Changes per Edge length",
      title = "Relative Transition Rates")


ggplot() + 
  geom_point(data = avg_trans_times %>% filter(transition == "armature->no_armature"), 
             aes(x = avg_time, y = avg_Nchanges / avg_edge.length, colour = "Armature -> no Armature"), 
             pch = 15) +
   geom_line(data = avg_trans_times %>% filter(transition == "armature->no_armature"),
             aes(x = avg_time, y = avg_Nchanges / avg_edge.length, colour = "Armature -> no Armature", group = N_sim)) +
  geom_point(data = avg_trans_times %>% filter(transition == "no_armature->armature"),
             aes(x = avg_time, y = avg_Nchanges / avg_edge.length, colour = "No Armature -> Armature"), 
             pch = 17) +
   geom_line(data = avg_trans_times %>% filter(transition == "no_armature->armature"),
             aes(x = avg_time, y = avg_Nchanges / avg_edge.length, colour = "No Armature -> Armature",group = N_sim)) +
  scale_color_manual(values=c("lightgrey", "#FFBB00"))+
  
  
    # Set plot aesthetics
    theme_light() + 
    xlim(45, 0) +
    ylim(0, 0.16) +
    labs(
      x = "Time since the origin of clade (mya)",
      y = "Mean Number of Changes/Mean Edge Length",
      title = "Mean Relative Transition Rates")

```

# B) Mammals

```{r}
#| label: loop
#| eval: false

rm(list=ls())
mtrees <- readRDS("../output/RDS/Mammals_4traits_simmap_100.rds")
tree <- readRDS("../input/Mammals/Processed/Mammals_MCC_drop.rds")
### Required function ========================================================================== #
getChanges<-function(tree){
  states<-sort(unique(getStates(tree)))
  nc<-sapply(tree$maps,length)-1
  b<-which(nc>0)
  nc<-nc[b]
  xx<-vector()
  H<-nodeHeights(tree)
  for(i in 1:length(b)){
    for(j in 1:nc[i]){
      ss<-names(tree$maps[[b[i]]])[j+1]
      x<-rep(H[b[i],1]+cumsum(tree$maps[[b[i]]])[j],2)
      xx<-c(xx,setNames(x[1],
                        paste(names(tree$maps[[b[i]]])[j:(j+1)],
                              collapse="->")))
    }
  }
  xx
}



# ============================================================================================= #
# Create a function to find the segment ID for a given time value:
# ============================================================================================= #

find_segment_id <- function(time) {
  segment_indices <- cut(time, breaks = segs_df[, 2], labels = FALSE)
  return(segs_df$segsID[segment_indices])
}

## ============================================================================================= #

```

```{r}
tic( "loop")

# ================================================================= #
# Working the multiSiammp
# ================================================================= #

# Get all changes from all simulations so that we can loop through them:
changes <- sapply(mtrees, getChanges) 
  
  
# Transform the original changes list of matrices from sapply() for all simulations to dataframe 
  # add trans_no to each transition based on grouping by transition 
  # (i.e, there are two trans_no = 1, one for each transition type)
  
  
i_changes_list <- list() 
for (i in 1:length(changes)) {
  i_changes <- changes[[i]] # for each simulation from multiSimmap
  
  # extract transition types and times and assign ID column for simmap simulation
  i_changes_df <- data.frame(transition = names(i_changes), time = i_changes, N_sim = i)
  i_changes_df$transition <- as.factor(i_changes_df$transition)
    
  # Add trans_no to each transition from a group (transition type)
  i_changes_df <- i_changes_df %>%  arrange(time) %>%
    group_by(transition) %>%
    mutate(trans_no = row_number())
    
  i_changes_list[[i]] <- i_changes_df
}
  
# Final Table without edge.lengths:
changes_table <- plyr::rbind.fill(i_changes_list, fill=T)
changes_table_wo_edge <- changes_table

write.csv(changes_table_wo_edge, "../output/CSV/4traits_Mammals_changes_table_wo_edgelengths.csv")



# ====================================================================================================== #
# ~ Step 1: Loop over each simulation and extract 
# - breaks for segments
# - edge.lengths
# - rates for each character state transition
# ====================================================================================================== #

object_list <- list()
rates_list <- list()

# First Level of the loop  
for (Nsimulation in 1:100){ 
  
  # ====================================================================================================== #
  # Setting global variables for the Analysis
  # ====================================================================================================== #
  
  # set the number of time-segments for which everything will be calculated:
  b <- 20*5    
  
  # reduce to one tree to extract parameters from each simulation separately:
  tree <- as.phylo(mtrees[[Nsimulation]])
  
  # extract time since root
  h <- max(nodeHeights(tree))
  
  # calculate the time segments based on time since the origin and the time segments that were set by the user
  segs <- cbind(seq(0,h-h/b,h/b),
                seq(1/b*h,h,h/b))

  
  # ====================================================================================================== #
  # Transitions & Timing  (getChanges)
  # ====================================================================================================== #
  
  
  # ================================================================= #
  # Working changes from each simmap simulation separately:
  # ----------------------------------------------------------------- #
  # Changes per Segments:
  # ================================================================= #
  


  # Loop through all changes & Calculate average number of changes per time interval (segs; set with "b")
  
  nchanges <- rep(0,b)
  for(i in 1:length(changes)) { # loop through all simulations from simmap  

    # loop through all individual changes from one simmap simulation and calculate relative changes per time segment
    # nchanges = collects the relative number of changes (of all changes) per segment
    
    for(j in 1:length(changes[[i]])) { 
      ind<-which((changes[[i]][j]>segs[,1])+ 
                 (changes[[i]][j]<=segs[,2])==2)
      nchanges[ind]<-nchanges[ind]+1/length(changes)
      
      } # closing of 2nd level
    } # closing of 1st level

  rm(i,j)  
  
  # ================================================================= #
  # Working changes from each simmap simulation separately:
  # ----------------------------------------------------------------- #
  # Edge Lengths per Segments:
  # ================================================================= #

  # Edge.Lengths = Number of species in the phylogeny per time slot (increases naturally with time)
  # ! Note: has to be accounted for when showing transition rates
  # Here we calculate edge lengths for each segement separatly:
  
  # -------------------------------------------------------------------------------------------------------- #
  

  
  # Compute Lineages-Through-Time with ltt() function from the phytools package to extract edge.lengths from
  LTT <- ltt(tree,plot=FALSE, log.lineages = F)
  cols<-setNames(c("#009E50", "#709E01", "#E69F00", "#E68F90", "darkgrey"),
                 c("herb_large","herb_small", "non_herb_large", "non_herb_small", "total"))

  plot(LTT,col=cols,bty="n",las=1,lwd=4,show.tree=FALSE,ylim=c(1,Ntip(tree)),
    xlab="time (above the root)")
  
  
	# transform LTT data so that we can use it:
  LTT <- cbind(LTT$ltt[2:(length(LTT$ltt)-1)],            # Number of species at each node minus the root
               LTT$times[2:(length(LTT$ltt)-1)],          # "start" time for each node minus the root
               LTT$times[3:length(LTT$ltt)])              # "end" time for each node minus the root
  
  # -------------------------------------------------------------------------------------------------#
  
  ## Now we loop through all segments to extract matching edge.lengths =================================== #
  
  # set count variable:
	ii<-1

	# Create an empty vector to store edge lengths
	edge.length <- rep(0, nrow(segs))

	
	# 2nd level loop: 
	# (for one tree because LTT is calculated for one tree)
   for (i_segs in 1:nrow(segs)) {
     

     # Find the indices of LTT segments that overlap with the current segs segment
     overlap_indices <- which(LTT[, 2] <= segs[i_segs, 2] & LTT[, 3] >= segs[i_segs, 1])
  

     # Calculate edge length for the current segs segment using vectorized operations
     # 3rd level loop: 
     for (ii in overlap_indices) {
     
       # edge.length should contain the edge lengths for all segments:  
       edge.length[i_segs] <- edge.length[i_segs] +        
         LTT[ii, 1] * # counts number of lineages between time intervals
         (min(segs[i_segs, 2], LTT[ii, 3]) - # end time interval
            max(segs[i_segs, 1], LTT[ii, 2]))  # start time interval

     
       } # 3rd level closing
     
   } # 2nd level closing
	
	
	edge.length_backup <- edge.length
	edge.length <-c(1, edge.length)
  # edge.length should contain the edge lengths for all segments in a given simulation
 	
	# adding segsID coulmn to edge lengths:
	edge.length_df <- data.frame(edge.length = edge.length, segsID = seq(1,b+1))
	
	

	# ================================================================= #
  # Working the multiSimmap:
  # ----------------------------------------------------------------- #
  # Edge Lengths at each transition event
  # ================================================================= #
	## Adapted code from the segments (above) by Frieda
	
	# we need the table for all simmaps: changes_table so that we can loop through it
	
	# set count variable:
	ii<-1

	# Create an empty vector to store edge lengths
	edge.length_transitions <- rep(0, nrow(changes_table))
  edge.lengths_all <- list()
	changes_times <- changes_table$time
	
	
	# 2nd level loop: 
	# (for one tree because LTT is calculated for one tree)
   for (i_changes in 1:nrow(changes_table)) {

     # Find the indices of LTT segments that overlap with the current segs segment
     
     # "which start times of the Lineages are smaller or equal to the time the change happened and which end times of the lineages are larger or equal to the time the change happend."
     # --> this gives the node numbers that were there when the transition happend
     
     overlap_indices <- which(LTT[, 2] <= changes_times[i_changes] & LTT[, 3] >= changes_times[i_changes])
  

     # Calculate edge length for the current segs segment using vectorized operations
     # 3rd level loop: 
     
     for (ii in overlap_indices) { # ii = number of lineages when a transition occurred
     
       # edge.length should contain the edge lengths for all segments:  
       edge.length_transitions[i_changes] <- edge.length_transitions[i_changes] +        
         LTT[ii, 1] 
       
       edge.length_transitions2 <- data.frame(time = changes_times[i_changes], edge.length = edge.length_transitions[i_changes])

     
       edge.lengths_all[[i_changes]] <- edge.length_transitions2
       
       } # 3rd level closing
     
   } # 2nd level closing
	
  
  edge.lengths_all_df <- plyr::rbind.fill(edge.lengths_all)
  changes_table_w_edge <- merge(changes_table_wo_edge, edge.lengths_all_df, by = "time")
  # Write to file (once is enough)
  write.csv(changes_table_w_edge, "../output/CSV/4traits_Mammals_changes_table_w_edgelengths.csv")
	

# ============================================================================================= #
# Create a dataframe with transitions, times, edge.lengths, segment IDs.
# ============================================================================================= #
	  
  ## We are here = first level loop: along simulations
	  
	 	# Again extract changes for each simulation separately:
	  current_simulation <- as.data.frame(changes[[Nsimulation]], 
	                                      row.names = names(changes[[Nsimulation]]))
	  names(current_simulation) <- c("time")
	  current_simulation$transition <- rownames(current_simulation)
	 	current_simulation$N_sim <- Nsimulation
	 	current_simulation <- current_simulation %>% arrange(time)
	 	current_simulation$IDtransition <- seq(1:nrow(current_simulation))

	 	
	 	# ============================================================================================= #
	 	# Working the segments: 
	 	# ============================================================================================= #

	 	
	 	# Create a vector of segment IDs
	 	segs_df <- as.data.frame(segs)
	 	new_row <- data.frame(V1 = 0, V2 = 0) # add a row that starts from 0 and ends with 0.. we need this later on for plotting
	 	segs_df <- rbind(new_row, segs_df)
	 	segs_df$segsID <- seq(1,b+1) # now we can calculate the breaks from it

	 	# Use the apply function to find the segment ID for each row in current_simulation
	 	current_simulation$segsID <- apply(current_simulation, 1, function(row) {
  
	 	  time <- as.numeric(row["time"])
	 	  return(find_segment_id(time))

	 	  }) # closing loop from within the apply-function (2nd level loop closing)

	
	 	# assign edge.length based on segment ID
	 	current_simulation <- unique(merge(current_simulation, edge.length_df, by="segsID", all = T))
	 	names(segs_df) <- c("segs_start", "segs_end", "segsID")
	  current_simulation <-	unique(merge(current_simulation, segs_df, by = "segsID", all = T))
	 	
	  # Calculate rate for each simulation:
	 	current_simulation_rates <- current_simulation %>% 
	 	  group_by(segsID, N_sim, transition) %>% 
	 	  mutate(Nchanges = length(!is.na(IDtransition))) %>% 
	 	  mutate(rate = Nchanges / edge.length)

	 rates_list[[Nsimulation]] <- current_simulation_rates # we need the rates_list object for plotting (!)
	  
	 object <-list(segments=segs, 
	               nchanges=nchanges, 
	               edge.length=edge.length[-1], 
	               tree=tree)
	 
	class(object)<-"ctt"
	object_list[[Nsimulation]] <- object 
	  
} # 1st level closing

toc()  # 1218.6 sec elapsed = 20.3 min

class(object_list) <-"multiCtt"

# -------------------------------------------------------------------------------------- #

# Make dataframe with rates from the rates_list object for all simulations from the simmap:
rates_df <- plyr::rbind.fill(rates_list)


# Now set columns: rates, Nchanges to 0 if there was no transition
# Calculate the average transition time based on numbered transitions from 100 simmap simulations
# set average timing of 0-transition to mean-segment time 
# then order the dataframe by segment ID:
rates_df <- rates_df %>%
  group_by(N_sim, segsID, transition) %>%
  
  mutate(rate = ifelse(is.na(transition), 0, rate), # set transition rates 0 if there was no transition
         
         Nchanges = ifelse(is.na(transition), 0, Nchanges), # set Nchanges 0 if there was no transition
         
         segs_mean = (segs_end+ segs_start)/2) %>% # Adjust the threshold (1e-6) as needed
  
  mutate(avg_time = ifelse(Nchanges == 1, time, mean(time))) %>%  # set average time = time if there was only 1 change
  
  mutate(avg_time = ifelse(is.na(avg_time), segs_mean, avg_time)) %>% # set time to mean time bin if there was no transition
  
  arrange(segsID) %>% unique() 


save.image("~/GitHub/Evo-Legumes-Herbivory/LegumesArmature/output/RData/4traits_Mammals_rates_ws.RData")
```

```{r}
rm(list=ls())
#load("../output/RData/Mammals_rates_ws.RData")

load("../output/RData/4traits_Mammals_rates_ws.RData")
```

## plot.multiCtt()

```{r}
#| label: plot.multiCtt() function Mammals

plot.multiCtt<-function(x,...){
  if(hasArg(alpha)) alpha<-list(...)$alpha
  else alpha<-0.05
  segments<-x[[1]]$segments
  nchanges<-sapply(x,function(x) x$nchanges)
  if(hasArg(type)) type<-list(...)$type
  else type<-"rate"
  edge.length<-sapply(x,function(x) x$edge.length)
  obj<-list(segments=segments,nchanges=rowMeans(nchanges),
            edge.length=rowMeans(edge.length),tree=x[[1]]$tree)
  class(obj)<-"ctt"
  lower<-max(floor(alpha/2*length(x)),1)
  upper<-min(ceiling((1-alpha/2)*length(x)),ncol(nchanges))
  xx<-max(nodeHeights(x[[1]]$tree))-as.vector(t(segments))
  xx<-c(xx,xx[length(xx):1])
  y.lower<-if(type=="number") apply(nchanges,1,sort)[lower,] else
    if(type=="rate") apply(nchanges/edge.length,1,sort)[lower,]
  y.upper<-if(type=="number") apply(nchanges,1,sort)[upper,] else
    if(type=="rate") apply(nchanges/edge.length,1,sort)[upper,]
  y.lower<-as.vector(rbind(y.lower,y.lower))
  y.upper<-as.vector(rbind(y.upper,y.upper))
  yy<-c(y.lower,y.upper[length(y.upper):1])
  args<-list(...)
  if(!is.null(args$alpha)) args$alpha<-NULL
  if(is.null(args$col)) args$col<-"blue"
  if(is.null(args$ylim)) args$ylim<-range(yy)
  args$x<-obj
  do.call(plot,args)
  polygon(xx,yy,col=make.transparent("grey",0.4),border=0)
}

plot.multiCtt(object_list, type="rate")
plot.multiCtt(object_list, type="number")
```

We can also plot the rates of each simulation separately using
plot.ctt() from phytools:

```{r}
#|label: plot.ctt() function

plot.ctt<-function(x,...){
  h<-max(nodeHeights(x$tree))
  args<-list(...)
  if(!is.null(args$type)){ 
    type<-args$type
    args$type<-NULL
  } else type<-"rate"
  if(!is.null(args$show.tree)){
    show.tree<-args$show.tree
    args$show.tree<-NULL
  } else show.tree<-FALSE
  if(!is.null(args$add)){ 
    add<-args$add
    args$add<-NULL
  } else add<-FALSE
  if(is.null(args$ylim)) 
    args$ylim<-if(type=="number")c(0,max(x$nchanges)) else 
      c(0,max(x$nchanges/x$edge.length))
  if(is.null(args$xlim))
    args$xlim<-c(max(x$segments),min(x$segments))
  if(is.null(args$lwd)) args$lwd<-2
  if(is.null(args$xlab)) args$xlab<-"time since the present"
  if(is.null(args$ylab)) 
    args$ylab<-if(type=="number") "mean number of changes" else
      "mean number of changes / total edge length"
  args$type<-"l"
  args$x<-h-as.vector(t(x$segments))
  args$y<-if(type=="number") rbind(x$nchanges,x$nchanges) else 
    rbind(x$nchanges/x$edge.length,x$nchanges/x$edge.length)
  if(!add) do.call(plot,args)
  else do.call(lines,args)
  if(show.tree) plotTree(x$tree,add=TRUE,ftype="off",lwd=1,
                         color=make.transparent("blue",0.1),mar=par()$mar,
                         direction="leftwards",xlim=args$xlim)
}


plot.ctt(object_list[[1]], type="rate")
plot.ctt(object_list[[1]], type="number")
```

# 2.1. Post-processing of data

***To do:***

Currently, if there was no transition of a specific type at a specific
time, there is no indication of a rate. However, the rate should be zero
if there was no transition. In the following code I try to replicate
each time segment twice and add transitions for both types with rates ==
0.\
I'm not done with this yet.

```{r}
# Make dataframe with rates from the rates_list object for all simulations from the simmap:
rates_df <- plyr::rbind.fill(rates_list)


# Now set columns: rates, Nchanges to 0 if there was no transition
# Calculate the average transition time based on numbered transitions from 100 simmap simulations
# set average timing of 0-transition to mean-segment time 
# then order the dataframe by segment ID:
rates_df <- rates_df %>%
  group_by(N_sim, segsID, transition) %>%
  
  mutate(rate = ifelse(is.na(transition), 0, rate), # set transition rates 0 if there was no transition
         
         Nchanges = ifelse(is.na(transition), 0, Nchanges), # set Nchanges 0 if there was no transition
         
         segs_mean = (segs_end+ segs_start)/2) %>% # Adjust the threshold (1e-6) as needed
  
  mutate(avg_time = ifelse(Nchanges == 1, time, mean(time))) %>%  # set average time = time if there was only 1 change
  
  mutate(avg_time = ifelse(is.na(avg_time), segs_mean, avg_time)) %>% # set time to mean time bin if there was no transition
  
  arrange(segsID) %>% unique() 

# Replace NAs in Transition types to set the rates to 0 where no transition occurred:
rates_df_copy <- rates_df

# Replace "NA" with the first transition type
rates_df_copy$transition[is.na(rates_df_copy$transition)] <- unique(rates_df_copy$transition)[[2]]

# Create a second copy of the data frame with "NA" transitions
rates_df_copy2 <- rates_df

# Replace "NA" with the second transition type
rates_df_copy2$transition[is.na(rates_df_copy2$transition)] <- unique(rates_df_copy2$transition)[[3]]

# Combine the two copies to get the final data frame with replicates
final_df <- bind_rows(rates_df_copy, rates_df_copy2)


########## ====================================================================================== ####
# Specify the transition type for which you want to ensure rates
specific_transition_type <- unique(rates_df_copy$transition)[[2]]

# Generate a data frame with all possible combinations of segsID and the specific transition type
all_combinations <- expand_grid(
  segsID = unique(final_df$segsID),
  transition = specific_transition_type
)

# Left join the generated data frame with the original data frame
final_df1 <- all_combinations %>%
  left_join(final_df, by = c("segsID", "transition")) %>%
  mutate(rate = ifelse(is.na(rate), 0, rate))# Generate a data frame with all possible combinations of segsID and the specific transition type
all_combinations <- expand_grid(
  segsID = unique(final_df1$segsID),
  transition = specific_transition_type
)

# Left join the generated data frame with the original data frame
final_df1 <- all_combinations %>%
  left_join(final_df1, by = c("segsID", "transition")) %>%
  mutate(rate = ifelse(is.na(rate), 0, rate))
########## ====================================================================================== ####
# Specify the transition type for which you want to ensure rates
specific_transition_type <- unique(rates_df_copy$transition)[[1]]

# Generate a data frame with all possible combinations of segsID and the specific transition type
all_combinations2 <- expand_grid(
  segsID = unique(final_df$segsID),
  transition = specific_transition_type
)

# Left join the generated data frame with the original data frame
final_df2 <- all_combinations2 %>%
  left_join(final_df, by = c("segsID", "transition")) %>%
  mutate(rate = ifelse(is.na(rate), 0, rate))# Generate a data frame with all possible combinations of segsID and the specific transition type

all_combinations2 <- expand_grid(
  segsID = unique(final_df2$segsID),
  transition = specific_transition_type
)

# Left join the generated data frame with the original data frame
final_df2 <- all_combinations2 %>%
  left_join(final_df2, by = c("segsID", "transition")) %>%
  mutate(rate = ifelse(is.na(rate), 0, rate))
########## ====================================================================================== ####



# Sort the data frame if needed
final_df <- final_df %>% arrange(segsID)  # You can specify the sorting order

# Now final_df contains rows with "NA" in the "transition" column replaced by two different transition types.










ggplot(final_df, aes(y = rate, x = segs_end, fill = factor(transition))) +
  geom_col(position = "identity")+
  ylim(0, 0.015)+
  theme_light()

```

### Plotting the rates

#### My adaptation of pollination study using

```{r}
## below adapted from other script:  https://github.com/rubysaltbush/pollination-macroevolution/blob/main/scripts/analysis/simmap.R


megaherbivore_transitions <- rates_df
megaherbivore_transitions$transition <- as.factor(megaherbivore_transitions$transition)


# build new data frame with cumulative number of transitions
megaherbivore_trans_cumul <- data.frame()
for(n in 1:100){
  trans <- megaherbivore_transitions %>% ungroup() %>%  
    dplyr::filter(N_sim == n) %>%
    dplyr::group_by(transition) %>% arrange(time) %>%
    dplyr::mutate(trans_no = row_number(N_sim))
  
  megaherbivore_trans_cumul <- rbind(megaherbivore_trans_cumul, trans)
}
rm(n, trans)




# =========================================================================================================== #
#  Now we are summarizing over the 100 simulations:
# =========================================================================================================== #
##### copied from somewhere else in the code from the same script....:


# first need to know average number of transitions and across simulations, rounded. 
trans_avg_length <- megaherbivore_trans_cumul %>%
  dplyr::group_by(transition, N_sim) %>%
  dplyr::mutate(no_trans = max(trans_no)) %>%
  dplyr::ungroup() %>%
  dplyr::filter(trans_no == no_trans) %>%
  dplyr::group_by(transition) %>%
  dplyr::mutate(avg_length = round(mean(no_trans), digits = 0)) %>%
  dplyr::select(transition, avg_length) %>%
  dplyr::ungroup() %>%
  dplyr::distinct()

# now knowing this, can rearrange data and average times and edge.lengths across rows
avg_trans_times <- megaherbivore_trans_cumul %>%
  dplyr::group_by(transition, trans_no) %>%
  dplyr::mutate(avg_time = mean(time)) %>%
  dplyr::mutate(SE_time = sqrt(var(time) / length(time)),
                avg_edge.length = mean(edge.length)) %>% ungroup() %>%
  group_by(N_sim, segsID, transition) %>%
  mutate(avg_Nchanges = mean(Nchanges)) %>%
  dplyr::select(transition, trans_no, avg_time, SE_time, avg_edge.length, avg_Nchanges) %>%
  dplyr::distinct()

# reduce avg_trans_times to trans_avg_length
avg_trans_times_m2o <- avg_trans_times %>%
  dplyr::filter(transition == "megaherbivore->other") %>%
  dplyr::filter(trans_no <= trans_avg_length[1,2])
avg_trans_times_o2m <- avg_trans_times %>%
  dplyr::filter(transition == "other->megaherbivore") %>%
  dplyr::filter(trans_no <= trans_avg_length[2,2])
avg_trans_times <- rbind(avg_trans_times_m2o, avg_trans_times_o2m)
rm(avg_trans_times_m2o, avg_trans_times_o2m)




#### Plotting ======================================

# install.packages("prettyGraphs") #to set transparancy of colors with "alpha"
library(prettyGraphs)

myColours = c("lightgrey", "#FFBB00")
myColoursAlpha <- add.alpha(myColours, alpha=0.4)
my_cols <- setNames(myColoursAlpha, c("megaherbivore->other", "other->megaherbivore"))


```

#### same thing in ggplot2 (rates plot)

```{r}
library(ggplot2)


plot.ctt(object, type="rate")

no_mega_to_mega <- megaherbivore_transitions %>%
      filter(transition == "other->megaherbivore")
    
mega_to_no_mega <- megaherbivore_transitions %>%
      filter(transition == "megaherbivore->other")


ggplot() +  
  # megaherbivore -> other:
      geom_point(data = mega_to_no_mega, aes(x = time, y = rate, colour = "megaherbivore -> other"), pch = 15, alpha = 0.3) +
   # other -> megaherbivore:
      geom_point(data = no_mega_to_mega, aes(x = time, y = rate, colour = "other -> megaherbivore"), pch = 17, alpha = 0.3) +
      geom_line(data = mega_to_no_mega, aes(x = time, y = rate, colour = "megaherbivore -> other", group = N_sim)) +
      geom_line(data = no_mega_to_mega, aes(x = time, y = rate, colour = "other -> megaherbivore",group = N_sim)) +
    # Set plot aesthetics
    theme_light() + 
    xlim(220, 0) +
    ylim(0, 0.1) +
  scale_color_manual( values=c("lightgrey", "#FFBB00"))+
    labs(
      x = "Time since the origin of clade (mya)",
      y = "Relative Changes per Edge length",
      title = "Relative Transition Rates")


ggplot() + 
  geom_point(data = avg_trans_times %>% filter(transition == "megaherbivore->other"), 
             aes(x = avg_time, y = avg_Nchanges / avg_edge.length, colour = "megaherbivore -> other"), 
             pch = 15) +
   geom_line(data = avg_trans_times %>% filter(transition == "megaherbivore->other"),
             aes(x = avg_time, y = avg_Nchanges / avg_edge.length, colour = "megaherbivore -> other", group = N_sim)) +
  geom_point(data = avg_trans_times %>% filter(transition == "other->megaherbivore"),
             aes(x = avg_time, y = avg_Nchanges / avg_edge.length, colour = "other -> megaherbivore"), 
             pch = 17) +
   geom_line(data = avg_trans_times %>% filter(transition == "other->megaherbivore"),
             aes(x = avg_time, y = avg_Nchanges / avg_edge.length, colour = "other -> megaherbivore",group = N_sim)) +
  scale_color_manual(values=c("lightgrey", "#FFBB00"))+
  
  
    # Set plot aesthetics
    theme_light() + 
    #xlim(220, 0) +
    ylim(0, 0.08) +
    labs(
      x = "Time since the origin of clade (mya)",
      y = "Mean Number of Changes/Mean Edge Length",
      title = "Mean Relative Transition Rates")

```
